A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-31 paper.tex
Light edits

The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
describe the regions of genome that have been passed from a parent to its immediate children,
hence edges either span the entire genome, or all the region to the left (or right) of a breakpoint
(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
it is inefficient to extract the local trees from the ARG structure.

@@

Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation schemes.
To obtain the sample-resolved annotation in Fig~\ref{fig-ancestry-resolution}B we must
traverse backwards in time from sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the
edge intervals to record regions of sample genome that have passed through each route.
When edges join into a parent node, the parent is allocated the union of the intervals.
When a child node is connected to two parents, one parent is allocated the edge
intervals of the child, truncated to include only genomic material to the left of the breakpoint.
Similarly the other parent is allocated the child's intervals truncated such that they contain
only regions to the right of the breakpoint (a natural extension exists for multiple
chromosomal breakpoints).

For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
transmitted immediately between parent and child: i.e. local to those two nodes in the tree.
However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
it now spans 0..2, the region that eventually ends up in
sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
derived locally: it must be done by traversing backwards in time from the sample nodes
up through the ARG.

@@

Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
the ancestral path from the samples it (and its associated edges) can be removed.
For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
substantially more efficient than immediate edge annotation. Second, by the time we reach
nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
multiple root nodes in different regions of the genome. Third, there exist nodes such as
\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
node \textsf{k}). in this case sample-resolved edge annotations lose the exact position of
the breakpoint, although they retain the information that a breakpoint occurred between
positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. \textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g. node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g. node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g. node \textsf{s}). These represent a ``stacking up'' of multiple
often indistinguishable events into a single node, something which is impossible
to encode in an eARG.

>>>>

The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
describe the regions of genome that have been passed from a parent to its immediate children,
hence edges either span the entire genome, or entire region to the left (or right)
of a breakpoint
(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
it is inefficient to extract the local trees from the ARG structure.

@@

Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation
schemes. To obtain the sample-resolved annotation in
Fig~\ref{fig-ancestry-resolution}B we must traverse backwards in time from
sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the edge intervals
to record regions of sample genomes that have passed through each route. When
edges join into a parent node, the parent is associated with the union of the
child intervals. When a child node is connected to two parents, one parent is
allocated the edge intervals of the child, truncated to include only genomic
material to the left of the breakpoint. Similarly the other parent is allocated
the child's intervals truncated such that they contain only regions to the
right of the breakpoint (a natural extension exists for multiple chromosomal
breakpoints).

For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
transmitted immediately between parent and child.
However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
it now spans 0..2, the region that eventually ends up in
sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
derived locally: it must be done by traversing backwards in time from the sample nodes
up through the ARG.

@@

Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
the ancestral path from the samples it (and its associated edges) can be removed.
For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
substantially more efficient than immediate edge annotation.
Second, by the time we reach
nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
multiple root nodes in different regions of the genome.
Third, there exist nodes such as
\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
the breakpoint, although they retain the information that a breakpoint occurred between
positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from any given individual tree.
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.
